---
title: "Upscaling the resilience assessment through comparative analysis"

abstract: |
  Resilience is the capacity of any system to maintain its function, structure and identity despite disturbances. Assessing resilience has been elusive due to high levels of abstraction that are difficult to empirically test, or the lack of high quality data required once appropriate proxies are identified. Most resilience assessments are limited to specific situation arenas, making comparision one of the unresolved challenges. Here we show how leveraging comparative analysis can provide insights on how Arctic communities (N = 40) can best deal with social and environmental change. We found that the capacity to self-organize, and nurturing diversity are sufficient conditions for Arctic communities whose livelihoods have been resilient, or for communities whose livelihoods have been transformed. Our study provides an alternative perspective on how to assess resilience by leveraging comparsion across cases. It also identify governance patways to support adaptations and transformations in the Arctic, a geography with some of the most dramatic social and natural challenges to come. 

bibliography: arctic_QCA.bib
csl: /Users/juanrocha/Documents/styles/elsevier-harvard.csl
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
    number_sections: true
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt

linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \linenumbers
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}

---

```{r data, include=FALSE}
library(tidyverse)
library(here)

## load datasets if it helps for writing
load(here("data", "cleaned_data.RData"))
load(here("data", "cleaned_df.RData"))
load(here("data", "coords.RData"))
load(here("data", "truth_tables.RData"))

```

# Introduction

Resilience is the capacity of a system to absorb disturbance and reorganize
while undergoing change so as to still retain essentially the same function,
structure, feedbacks, and therefore identity [@Anonymous:2016fv]. There are
over 70 definitions of resilience [@Fisher:2015fj], making it a hard concept
to grasp, to operationalize, and empirically ground. How does one measure
resilience remains a non-trivial challenge and a key frontier of active research
[@Liu:2020ku; @Gao:2016dn; @Scheffer:2009p4449].

Foundational work on resilience emerged from reflections about ecological
stability [@Lewontin:1969vh] and the possibility that ecosystems can undergo
regime shifts [@Holling:1973p6861]. Regime shifts are abrupt and persistent
changes in the function and structure of systems [@Anonymous:2004bq]. Thus if
regime shifts relate to the possibility that a system can exist in multiple
configurations, resilience is the ability to keep a system within a particular
configuration. Regime shifts and resilience are two complementary sides of the
same coin. Not surprisingly, early theoretical work attempting to operationalize
resilience focused on metrics such as the size of the basin of attraction, its
depth, slope, or proximity to the boundary that separate different regimes
[@Clark:1975vj]. Recent efforts to operationalize resilience that build on
these approaches suffer from the problem of finding sufficient observables
[@Liu:2013gg], or high quality multidimensional data to characterize
configurations of social-ecological systems [@Dakos:2015jt].

While resilience is often introduced using ecological problems, its roots are
diverse. The use of the term in the social sciences precedes its use in ecology,
and the concept of regime shifts would have not been possible without the
development of bifurcation theory in mathematics almost a century and a half ago
[@Poincare:1885ji]. Bifurcation theory was applied to develop theories of
poverty traps in the 1950s [@Bowles:2006hx], segregation in sociology in the
1970s [@Schelling:1969wm], and over the last half century applied to fields
that span medicine, climate science, linguistics, finance, physics
[@Sole:2011us], and social-ecological systems [@Scheffer:2009wl;
@Anonymous:2004bq].

If resilience and related concepts are so widely used, why is it so hard to find
good empirical proxies of resilience? One barrier to the general applicability
of resilience is that it requires answering *a priori* the questions of
resilience of what to what? [@Carpenter:2014fw]. Defining resilience to
specific contexts reduces the ability to compare across cases. For example, one
might successfully approximate the resilience of a cancer, but it does not mean
one can use the same method or proxies to assess the resilience of a firm, or
even a different type of cancer. Consequently, most approaches to resilience
assessments have been centered around particular problems by particular actors
(e.g. water management), and remain participatory consultations limited by the
perceptions of participants. They are important exercises to generate shared
knowledge of a management problem that enable participants to identify
feedbacks, slow variables or potential leverage points for intervention
[@ResilienceAlliance:2011vc; @Walker:2012wp; @Walker:2012ur;
@sellberg2015resilience; @enfors2021wayfinder]. However, quicker, more generic,
cross-system approaches are needed [@Scheffer:2012cta; @Scheffer:2003p339;
@Scheffer:2009p4449].

Time being a common dimension to any dynamical system lends itself for the
development of generic proxies of resilience [@Scheffer:2009p4449], in
particular recovery time [@Holling:1973p6861]. Complex systems when close to
regime shifts leave statistical signatures on time series that reveals
resilience loss [@Strogatz:2014wo; @Scheffer:2012cta; @Scheffer:2009p4449],
in a nutshell it takes longer to recover after small disturbances
[@Scheffer:2009p4449; @Dakos:2015jt]. For example, in a person the time it
takes to recover after a stressful event; or in a forest, the time it takes to
recover after a series of fires. These methods can fail when dynamics are driven
by stochastic processes [@Hastings:2018gy; @Hastings:2010p5336], and often
require long time series to detect meaningful signals [@Dakos:2015jt].
Time-based metrics offer no insight into two important evolutionary dimensions
of resilience: adaptive and transformative capacities [@Anonymous:2016fv].

Another common approach to assessing resilience has been to develop frameworks
that assess the state of a variety of system variables that are believed to be
related to resilience. Resilience researchers and practitioners have developed a
wide variety of different resilience assessment guides and frameworks
[@sharifi2016critical]. A social-ecological example of such a framework is
Toolkit for the Indicators of Resilience in Social-Ecological Production
Landscapes and Seascapes, which asks its users to score 20 indicators of
resilience with a score from 1-5 indicating minimal support to resilience to a
lot of support of resilience [@bergamini2013indicators]. The developers of this
toolkit have argued that when used as part of a community process this toolkit
can help its users understand and identify elements of socio-ecological
resilience, challenges to resilience, and potential strategies to address their
situation [@dunbar2020lessons]. While these frameworks are often a rapid way of
making sense of a complex situation, most of these frameworks have not been
tested, and most are not clearly linked to theory [@quinlan2016measuring].
Theory based frameworks make it easier to translate findings across cases, and
make it easier for practice to inform theory. Resilience frameworks that are
based on theory make it easier to translate between frameworks and to compare
cases.

Here we propose and test a qualitative comparative method for upscaling
resilience assessments. We study $`r nrow(coords)`$ communities in the Arctic
facing environmental and social challenges related to potential regime shifts in
Arctic ecosystems, or socio-economic and political pressures opened by recent
pathways of development (Fig \ref{fig:stats}). We outline how leveraging the
power of comparison enables extracting patterns that transcend individual case
studies informing theory development, as well as potential management or
governance arrangements. We hope our qualitative proposal complements current
quantitative metrics that focus on recovery time by gaining insights about the
ability to adapt and transform that define communities in the Arctic. But why
the Arctic?

The Arctic is a region of the world where the resilience of communities is being
tested by rapid climate, social and ecological change; but also a region with a
wealth of case studies that enable a comparative approach. Climate change is
impacting Arctic ecosystems \~4 times faster than the rest of the globe
[@Ford:2015hq]. Climate impact is greater than ice loss, it is fundamentally
changing landscape and seascape connectivity, the flow of energy and nutrients,
and structure and function of biological communities. The Arctic is also the
scene of vast socio-economic changes and challenges that are occuring in a
region that has been strongly shaped by the colonialism of Arctic nations
[@Carson:2016ts]. The reduction of Arctic sea ice is opening the possibility
for mining exploration and new international shipping routes
[@Barnhart:2015cb; @Huntington:2015fx], increase in pollution
[@Schroder:2017cq; @Law:2017cs] or changing conditions for fisheries and
aquaculture [@Troell:2017di]. Securing basic food, water and energy, as well
as diversifying economies, remain key human development challenges for Arctic
communities [@Larsen:2015bj].

How can Arctic communities respond to these social and ecological challenges?
Will they adapt and continue with their traditional livelihoods, or transform
towards new social-ecological configurations?

# Methods

To answer these questions we first revisited the literature on adaptive capacity
and designed a coding scheme for capturing elements of resilience across case
studies. We co-designed and co-developed a template to guide case study
selection, literature reviews, and validation with experts. Then we applied
qualitative comparative analysis to infer necessary and sufficient conditions
for resilience, transformations or resilience loss in $`r nrow(coords)`$
communities across the Arctic. Below we describe in detail each of these steps.


**Measuring adaptive capacity**: We draw inspiration from early conceptual
developments on how to approximate adaptive capacity. Adaptability is related to
the capacity to learn, combine experience and knowledge, innovate, and adjust
responses and institutions to changing external drivers and internal processes
[@Anonymous:2016fv]. In an influencial synthesis of research on resilience in
social-ecological systems, Berkes et al [-@Berkes:2003wx] proposed a two-tier
set of variables grouped around four key dimensions: i) learning to live with
change and uncertainty (N), ii) nurturing diversity for reorganization and
renewal (D), iii) combining different types of knowledge for learning (K), and
iv) creating opportunities for self organization (S). We operationalized the
two-tier approach with a third-tier of variables (Fig \ref{fig:vars}) by
combining recent developments on principles for building resilience
[@Anonymous:2011iw] with expert knowledge elicitation. The refining of our
third-tier of variables benefited from input and testing of preliminary case
studies in the context of the Arctic Resilience Assessment [@Carson:2016ts].


The Arctic Resilience Assessment was commissioned by the Arctic council under
the Swedish Ministry of Environment chairmanship
[@Carson:2016ts]. It builds on collaborations between the Arctic countries,
Indigenous peoples in the region, and several observers that include NGOs and
academic institutions. The research questions, methods development, and
selection of cases were informed and shaped by representatives of member states
and permanent participants representing Canada, Denmark, Finland, Iceland,
Norway, the Russian Federation, Sweden, the Aleut International Association, the
Arctic Athabaskan Council, the Gwich'in Council International, the Inuit
Circumpolar Council, the Russian Association of Indigenous Peoples of the North,
Siberia and Far East, and the Saami Council. We also received input from
representatives of the Arctic Council working groups of the Arctic Contaminants
Action Program, the Arctic Monitoring and Assessment Programme, Conservation of
Arctic Flora and Fauna, Protection of the Arctic marine Environment, and the
Sustainable Development working group of the Arctic Council. We also benefited
from contributions from Arctic Council observers, who included participants from
the European Environmental Agency, the International Arctic Science Committee,
the International Arctic Social Science Association, the Arctic Centre -
University of Lapland, the University of the Arctic, the Resilience Alliance,
the Stockholm Environmental Institute, and the WWF Global Arctic Programme.

**Co-designing a case study template**: We initially designed a template to
capture social-ecological dynamics from literature reviews and expert
elicitation inspired by the Regime Shifts database [@Biggs:2018hx] available
at: [www.regimeshifts.org](www.regimeshifts.org). The draft template was
modified to incorporate the feedback, needs and priorities of the different
actors involved in the Arctic Resilience Assessment. For example, we divided
sources of social and ecological resilience separately, we left Arctic regime
shift information optional and focus the template on timelines and historical
events that have challenged the communities studied. We modified and tested the
template through workshops and focus groups that took place in Washington,
Tromsø, Bodø, and Stockholm from 2015-2016. These workshops were used to present
the template and receive feedback from stakeholders, as well as test it with
particular case studies co-developed with case experts. They recommended us
literature, we reviewed and filled up the template, and then we checked with
them if the template has captured well enough the key features from the case as
well as the correct coding of adaptive capacity features. The group also guided
our selection of case studies to prioritize their most acute needs, and secure
an heterogeneous selection of studies across problem issues and geographies.
Specifically, we aimed to collect multiple cases from all Arctic Council
countries, which included key livelihoods in the region, and we aimed to equally
represent cases which were believed to have experienced loss of resilience,
resilience, or transformation.

The template was designed to capture key aspects of resilience and adaptive
capacity based on literature reviews. Cases were classified as having
experienced loss of resilience, resilience, or transformation, depending on the
assessment of case literature and case experts. Experts suggested cases that
they believed had been researched and had experienced one of these outcomes. A
social-ecological system was considered resilient if the main livelihoods of the
communities described are maintained over time, despite social and ecological
disturbances. A case is coded as transformation if a community shifts their main
source of livelihoods to a different set, but manages to continue developing
within the same place. A case is coded as loss of resilience where the main
livelihoods are lost, no new livelihoods are established, and symptoms of
community collapse long last such as emigration, simplified local economies,
high unemployment or high substance abuse (Fig \ref{fig:map}). In some of our
cases, other experts may define the system differently and consider different
livelihoods, therefore while our cases have geographic names we also state which
key issue has been used to define each social-ecological system for our
analysis. We attempted to have all case templates validated by case experts, who
were involved in the Arctic Resilience Assessment. These experts provided
suggestions of other literature, connections to other experts, and clarified our
interpretation of the case. All templates collected are available in
([10.6084/m9.figshare.14779533](doi.org/10.6084/m9.figshare.14779533)) data
repository. Table 1 presents a summary of the case studies.


**Qualitative comparative analysis**: Based on our case study templates, we
created a matrix of cases versus the third-tier adaptive capacity proxy
variables (Fig \ref{fig:vars}). Each entry is coded as -1 if the evidence in the
literature or expert elicitation suggest the absence of that variable, 1 if
evidence suggests presence of the variable, or 0 if there is no evidence
available. We aggregated the scores at the first-tier level by summing across
the third-level categories. Qualitative comparative analysis was then used to
investigate necessary and sufficient conditions for a case to be resilient,
transformation, or loss of resilience. A condition is necessary if it must be
present for a specific outcome to occur, while a condition is sufficient if it
can produce a certain outcome by itself [@Ragin:2009te].

Qualitative comparative analysis (QCA) is a method that uses boolean algebra
[@Boole:1854ww] and (fuzzy) set theory to investigate research questions with
an intermediate number of cases [@Ragin:2009te]. While QCA was developed in
political science and sociology [@Rihoux:2013jx], the method has been
previously applied in social-ecological systems research [@Crona:2015ii;
@Anonymous:2013iw; @Basurto:2013ev]. In our analysis, the four dimensions of
adaptive capacity [@Berkes:2003wx] were used as conditions, and the outputs
were coded as loss or resilience, resilience, or transformation. We conducted
QCA on the case matrix through a two step process.

First, we calibrated thresholds for inclusion or exclusion in a bolean set using
a logistic function where values lower than -2 were excluded, +2 included, and
-0.5 was assumed to be maximum ambiguity. For example, if a case has a score \>2
for variables under the category "*nurturing diversity for reorganization and
renewal*", the case is set to belong to the set **D**, which is a shorthand for
diversity; if the score was \<2, it does not belong to the set, which in QCA is
denoted as **d**. Throughout the paper, capital letters are used to denote
belonging to a set or presence of a feature, while small caps used for its
absence or negation.

Next, we computed subsets of necessary and sufficient conditions for each
possible output (i.e. loss of resilience **L**, resilience **R**, and
transformation **T**) and its negation (i.e. **l**, **r**, and **t**
respectively). We used the truth table of our case studies --a cross tabulation
of the number of cases in our sample against all possible combinations of
conditions and outputs-- to perform boolean minimization and identify
parsimonious, complex, and intermediate solutions. These solutions (or causal
recipes in the language of QCA) suggest a combination of conditions that leads
to an output (being resilient) to be true or false. In other words, it helps us
identify which features increase adaptive capacity, and hence resilience, in the
Arctic communities in our study. The analysis was performed in `R`
[@RCoreTeam:2012wf] using the `QCA` package [@Thiem:2012dq;
@Dusa:2019gp]. The code used in our analysis is available at:
<https://github.com/juanrocha/Arctic_QCA>

# Results

```{r echo = FALSE}
stats <- coords %>%
  select(Climate:`Other socio-economic`) %>%
  pivot_longer(1:last_col(), names_to = "driver", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(driver) %>%
  summarize(n = sum(value))

countries <- coords %>%
  select(country) %>%
  table()
```

The $`r nrow(coords)`$ cases analysed showcase some of the challenges that Arctic
communities are currently facing (Fig \ref{fig:stats}). The sectors most
impacted across countries are fisheries and hunting, both for commercial and
subsistence purposes. The most common challenges across our case studies are
climate ($`r stats %>% filter(driver == "Climate") %>% pull(n)`$ cases),
geopolitical conflicts
($`r stats %>% filter(driver == "Geopolitical") %>% pull(n)`$), mining
($`r stats %>% filter(driver == "Mineral extraction") %>% pull(n)`$), and other
socio-economic drivers
($`r stats %>% filter(driver == "Other socio-economic") %>% pull(n)`$). The
country with most cases in our dataset is Canada ($`r countries["Canada"]`$),
while Sweden, Finland and Iceland have only two cases each. While most of our
case studies describe relatively recent phenomena, one of them focuses on the
19th and 20th century, and another goes back to the 1600's. The time span
reported in each case study depends on the literature reviewed.

We tested different approaches for calibrating the data and found that a
logistic function worked best (Fig. \ref{fig:cal}). Alternative approaches such
as the empirical cumulative distribution function, or the totally fuzzy relative
method, use the distribution features of the underlying variable to set
thresholds for the inclusion or exclusion to a set [@Thiem:2012dq;
@Dusa:2019gp]. These data driven approaches are agnostic to the meaning of
variable values. However, in this study, the zero values have a specific
meaning, indicating features for which there is maximum ambiguity (Fig
\ref{fig:cal}). The logistic function allows us to incorporate that knowledge
into the calibration of fuzzy sets [@Dusa:2019gp] by setting thresholds for
inclusion and exclusion as (2, -2) respectively, and maximum ambiguity at -0.5.
The wider the inclusion / exclusion threshold ($>|2|$), the more fuzzy the
calibration is. At the limit, where the inclusion, exclusion, and ambiguity
thresholds are the same, the data become crisp sets (i.e. 0 or 1). Our analysis
is done with fuzzy sets because it utilizes the diversity of features to cover
many of the possible combinations of cases. A crisp set analysis does not
converge given poor coverage in the truth table.

The truth table reveals that our $`r nrow(coords)`$ Arctic cases capture
`r tt_list[[1]]$tt %>% filter(OUT != "?") %>% nrow()` out of the $`r 2^4`$
possible combinations of adaptive capacity features (Fig. \ref{fig:ttable}). In
qualitative comparative analysis, a case can either belong or not to a set.
Therefore, with four features of adaptive capacity [@Berkes:2003wx], there
are $2^4$ possible combinations of features. Despite having over twice (N =
$`r nrow(coords)`$) as many cases as possible combinations (16), our coverage is
less than half. Our cases only exhibit only 9 of the 16 possible combinations of
variables, and cases unevenly exhibit these combinations (e.g. there are 10
cases which have the logical combination **NDKS** while no cases are **ndkS** or
**ndKs**). This phenomena of limited diversity is common in the social sciences
[@Dusa:2019gp; @Ragin:2009te], and QCA uses the unobserved combinations
(logical reminders) to optimize different types of boolean minimization. For
example, the logical combination of features **nDks** or **ndKs** are cases that
could be classified simultaneously as belonging to the sets of resilient,
transformation, or loss of resilience (Fig \ref{fig:ttable}B). However, the
sufficiency inclusion score (incl) and the proportional reduction of
inconsistency (PRI) suggest that such cases are more likely to belong to the
loss of resilience set.

The analysis of necessity and sufficiency reveals that set of adaptive capacity
features for resilient and transformation cases are very similar. Both have the
same sufficient conditions with relatively similar scores for inclusion and
consistency reduction (Fig \ref{fig:necsuf}A.). While many of our cases lack the
capacity to self-organize (S), the highest scores for inconsistency reduction is
attributed to cases that rank high on self-organization (S), even in the absence
of other attributes such as the ability to navigate uncertainty (N) or nurturing
diversity (D). The analysis of necessity does not clearly distinguish the three
possible outcomes (resilience, transformation, or loss of resilience) because
the ability to navigate change and uncertainty (N), nurturing diversity (D), or
combining different types of knowledge (K) are necessary conditions for all
outcomes.

We used Boolean minimization to discover what combinations of adaptive capacity
features provide parsimonious solutions, or a minimum set of conditions, for
each outcome (resilience, transformation, loss of resilience; Fig
\ref{fig:sols}). This analysis also showed that the adaptive capacity features
of resilience and transformation cases are very similar, where the capacity to
self-organize (S) appears to be a key feature. Resilient cases need to have at
least two out of the four adaptive capacity features, and the strongest
contributions are for cases that include $d*K*S$ or $N*K$. Transformation cases
follow a similar pattern, where most cases require at least two out of the four
adaptive capacity dimensions, with the most important appearing to be
self-organization (S), followed by a combination of knowledge (K) and the
capacity to navigate uncertainty (N). Cases with loss of resilience commonly
lack 3 out of the 4 features ($n*k*s$), even in the presence of different types
of knowledge (N) and the capacity to self-organize (S).


# Discussion

On summary, our analysis shows that all features of adaptive capacity are
important for resilience and transformation, but the capacity to self-organize
(S) and nurturing diversity of social and ecological sources of memory (D)
appear to play a larger role that the other features. Many of the Arctic cases
studied are showing symptoms of resilience loss and no sign of transformations
or shifting to alternative livelihoods (N =
$`r coords %>% filter(output == 0) %>% tally() %>% pull()`$). Communities that are
coping today are expected to face more disruptive change in the near future.
Therefore, the Arctic Council is interested in what management and governance
interventions could help communities enhance their adaptive capacities. For
example, nations could strengthen the ability of communities to manage their
territories, having the autonomy to craft their own rules, and getting support
for long term monitoring or a legal framework that promotes autonomy. Another
example is nurturing networks of knowledge exchange where people can share
success and failure stories for tackling similar issues, from climate change
impacts, to herd management, or dealing with novel economic activities that can
be in conflict with traditional use of their territories. Some such platforms
already exist, such as the International Centre for Reindeer Husbandry and the
Arctic Council itself, but there is a great need to enable such learning
networks across the Arctic.

Qualitative comparative analysis as a method proved useful to analyze a
relatively small sample size against an enormous combinatorial space of
possibilities. If the analysis was run at the second-tier variables (N =
$`r dat %>% pull(x2) %>% unique() %>% length()`$), one would have $2^{13}=$
$`r 2^13`$ combinations, and one would need a similar amount of cases. Our
solution to simplify from the third-tier proposition (N =
$`r dat %>% pull(x3) %>% unique() %>% length()`$) to the four features of adaptive
capacity was to add across each of them. The aggregation by summing has the
implicit assumption that all categories have equal importance. That is, the
researcher does not have an *a priori* information to decide which variables
could be more relevant for adaptive capacity. Our analysis presents a first
order approximation where no assumption of importance is made, nor bias is
induced to particular features of adaptive capacity. This assumption could be
changed by using other aggregation schemes (e.g. geometrically weighted) or
motivating a different calibration for the QCA, as theory develops on how
different social-ecological features interact to influence the resilience of
cases.

In our QCA analysis, we tested different Boolean minimization techniques. We
only report the parsimonious solutions, because when we used intermediate and
complex solutions they did offer different results. Intermediate solutions are
tested when the logical reminders (logical combinations for which we do not
observe cases, Fig \ref{fig:ttable}A) are assumed to be true. Because we have a
high proportion of combinations that were not observed (7/16), this assumption
does not simplify our results any further. Complex solutions are computed by
explicitly coding theoretically based expectations, for example that being
strong on the four features of adaptive capacity would lead to a resilient
outcome ($NDKS \rightarrow R$). However, this approach was not useful because
many of the cases that are coded as resilient or transformation lack at least
one of the adaptive capacity features (Fig \ref{fig:ttable}B).

This study develops a theory based approach to assess adaptive capacity. We
operationalized Berkes et al [-@Berkes:2003wx] by developing by developing a
third-tier of variables that can be coded from literature reviews. We developed
these variables because we initially failed when attempting coding cases at the
second-tier proposed by Berkes et al [@Berkes:2003wx], or the recently
developed principles for building resilience [@Anonymous:2011iw]. These
sources, while conceptually appealing, are too general to be able to code
presence or absence of features at the case study level. These concepts are
strongly rooted in the scholarly community of social-ecological systems and a
fair amount of conceptual translation was required on our part to share the
template and develop categories that were codable with and by other scholars, or
validation with stakeholders during the process. We hope that the effort we have
put into this study can be used by others to develop more accurate and effective
methods to rapidly assess the potential for transformation, resilience, and loss
of resilience in social-ecological systems.

Our study is based on published literature and expert elicitation, and as such
it is limited by their focus and scope. For example, several of our cases come
from ethnographic work with fishing and hunting communities which typically
involve male informants. Thus, our cases might lack other analytic dimensions
such as inequality or gender that the literature reviewed was not necessarily
targeting. Our study does have to some degree internal validation in the sense
that experts on the ground have validated our interpretation of literature and
coding of cases. However, external validation is not granted in the sense that
our results might not generalize to other social-ecological systems or contexts.
We encourage the academic community to test the external validation of our
results in other social-ecological systems.

Additionally, comparing our assessment of adaptive capacity to resilience
outcomes, reveals some ways to potentially improve our assessment of resilience.
The adaptive capacity features we used are based on social-ecological research
that de-emphasizes flows of energy, resources, and money that characterize urban
environments. For example, the resilient case that scores lowest in all adaptive
capacity features is Kiruna, an iron mining town in Northern Sweden, which is
being moved [@Nilsson:2010jf]. The Swedish government owned mining company in
Kiruna is spending over a billion Euros to move the town in order to expand the
large and profitable mine [@Nilsson:2010jf]. Kiruna's resilience is strongly
tied to an international company, global iron markets and a huge deposit of
iron, but these features are not included in our assessment of adaptive
capacity. This mismatch highlights the need to better incorporate cross-scale
interactions in resilience assessment [@sharifi2016critical;
@sellberg2015resilience]. Further work on assessing resilience should target
these limitations by integrating theory with new sources of data and multi-scale
resilience assessment processes.

In summary, this paper presents three contributions. First, it proposes
comparative analysis as a way to empirically assess the resilience of a diverse
set of communities in the changing Arctic. Second, it further develops the
adaptive capacity framework with a third tier of variables that enable us to
capture nuances where the second tier or alternative frameworks failed. Third,
the comparative approach allowed us to make some generalizations on common
adaptive capacity features that characterize cases where livelihoods were
resilient or transformed. These features highlight the importance of the
capacity to self-organize and combine knowledge diversity. Future research can
focus efforts in refining the third tier here proposed, exploring a weighted
approach that takes into account variable importance, comparing cases in other
geographies, or expanding our framework to time varying QCA. The latter would
help explore how the adaptive capacity features change over time.

# Conclusion

Resilience is a fundamental feature of complex social-ecological systems. Yet,
approaches for measuring resilience, adaptive or transformative capacities
remain confined to single case studies, limiting theory testing, hypothesis
development, or simply comparison. Here we developed a qualitative approach for
measuring resilience inspired by the adaptive capacity framework
[@Berkes:2003wx ]. Through a co-production process, we expanded the framework
with a third tier variable, and used comparative analysis to identify features
that increase resilience in Arctic cases. We found that cases with high capacity
to self-organized, to combine multiple sources of knowledge, or who nurture
diversity are more likely to maintain their livelihoods or transform towards new
social-ecological configurations. We believe our methodological approach can
complement existing quantitative methods such as early warning signals to
approximate resilience and participatory reslience assessment approaches. We
encourage researchers developing resilience frameworks to link their frameworks
to resilience theory and compare results across cases. We hope this work will
encourage resilience researchers to test and improve our coding approximation in
further social-ecological analyses.

# Box 1. Case examples

### Resilience: Pangnirtung- Nunavut Territory-Canada: Individual and collective adaptive strategies to face climate change

Located on Baffin Island, Pangnirtung is an Inuit fishing community that has
developed collective and individual adaptive strategies to respond to ecological
and social-economical disturbance. Sea-ice conditions, landscape, seascape,
weather conditions are the main effect of climate change in the area. Meanwhile,
socio-economic changes are related to change in Inuit fishing, lifestyle and the
market. To cope with these social ecological changes, communities have mixed
economies having commercial fisheries owned by Inuit people and subsistence
fisheries as their traditional livelihoods. The coexistence of these
livelihoods, among the introduction of technology, sharing networks, social
learning, strong indigenous institutions and the integration of different
knowledge systems has given the community the flexibility to deal with
uncertainty under climate change and social-economic transformations.

### Transformation: Húsavík in Skjálfandi Bay- Iceland: Fishing to whale watching tourism -- Iceland

Iceland's industry has historically been dominated by fishing. However, over the
last few decades, changes in international regulations affected the productivity
of the fishing sector. Employment in fishing communities declined due to reduced
cod quotas and the introduction of tradable quotas, which led to the fishing
industry consolidating into a few larger companies. Fishing also decreased due
to the international whaling moratorium, which also led to increases in whale
populations. These changes stimulated a reorganization of Húsavík's fishing
community to whale-watching tourism, based on a new application of existing
knowledge and skills. Today, Húsavík is one of the main whale-watching hotspots
in Iceland. Several other Icelandic fishing communities have made a similar
transition.

### Resilience loss: Aklavik- Northwest Territory-Canada: Human-whale relation in environmental and social changes.

Aklavik is a settlement from Northwest Territory, inhabited mainly by Inuvialuit
people. Residents used to hunt beluga as the primary food source and as a key
element of their cultural identity. There has been a decrease in these
practices, risking losing the knowledge and cooperative value of shared food
systems, and collaborative hunting. Climate change has impacted the sea ice
dynamics of the area and the weather stability, decreasing the places available
to hunt whales and increasing the risks of hunting. The integration of Western
values, technology, time in formal education, wage labor, external food supplies
and climate effect has demotivated young people to hunt. Among these, the aging
of elders is risking knowledge and skill transmission. However, communities
identify as hunters and recognize the importance of protecting this practice to
adapt.

# References

::: {#refs}
:::

\pagebreak

```{r echo = FALSE, results = "asis"}
library(knitr)
coords %>%
  arrange(desc(output)) %>%
  select(1:3) %>%
  kable(., caption = "**Case studies**. Their outcome are coded as resilient [R], transformation [T], or loss of resilience [L].", label = "tab1")

```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{fig1_arctic_map}
\caption{\textbf{Arctic case studies} $`r nrow(coords)`$ cases were analyzed across Arctic nations.}
\label{fig:map}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 3.5in]{fig2_stats}
\caption{\textbf{Descriptive statistics} $`r nrow(coords)`$ cases were analyzed across Arctic nations (A) covering a range of sectors (B) and responding to different types of drivers (C). The time spam of our case studies depend on what is reported on literature reviews. While most concentrate on contemporary issues, some cases date back centuries (D). The category "other" in (A) refers to the case of the Bering Strait which lies in international waters between Russia and USA.}
\label{fig:stats}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4in]{vars_dendrogram}
\caption{\textbf{Adaptive capacity} We operationalized the original proposition by Berkes et al to include a third tier of variables that enable the identification of adaptive capacity features from literature reviews and expert elicitation. The variables were co-designed and tested with a broad range of stakeholders.}
\label{fig:vars}
\end{figure*}
```


```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 3.5in]{fig3_calibration_heatmap}
\caption{\textbf{Calibrating adaptive capacity} Scores were calculated by adding all variables grouped by first-tier categories: i) learning to live with change and uncertainty (N), ii) nurturing diversity for reorganization and renewal (D), iii) combining different types of knowledge for learning (K), and iv) creating opportunities for self organization (S). Cases and adaptive capacity features are clustered around their scores, showing groups of resilient and transformation cases in the top, and resilience loss in the bottom (A). A logistic function was used to calibrate the raw data (B) for input into the qualitative comparative analysis. Fig \ref{fig:sm2} provides the raw values for the second-tier adaptive capacity features.}
\label{fig:cal}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{fig4_truth_tables}
\caption{\textbf{Truth tables} The truth tables show the distribution of cases across logical combinations. With 4 adaptive capacity features, there are 16 plausible combinations of these features. Our cases only cover 7 of them (A). The sufficiency inclusion score (incl) indicates how much a logical combination is sufficient for an outcome (e.g. OUT Resilience = True), while the proportion of reduction in inconsistency (PRI) reveals potential contradictions with respect to all other combinations.}
\label{fig:ttable}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{fig5_necessary_and_sufficient}
\caption{\textbf{Necessary and sufficient conditions} The analysis of (A) sufficiency and (B) necessity compares the relevance of observed combinations of features for an output. For sufficiency these combinations of features are multiplicative (set intersection, logical AND); while for necessity, these combinations of features are additive (set union, logical OR). The coverage (cov) and inclusion (incl) statistics help comparing the contribution of features to the outcome, while the proportional reduction of inconsistency (PRI) and relevance of necessity (RoN) weigh the relative contribution of each solution with respect to the other solutions.}
\label{fig:necsuf}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{fig6_solutions}
\caption{\textbf{Parsimonious solutions} Minimal set of conditions (rows) that can lead to different types of outcomes (resilience, transformation, or loss of resilience). Each term is a set of sufficient conditions that combined lead to a full minimal parsimonious solution. For example, the outcome resilience can be reached when a case  $n*k*s + N*K + D*s + d*K*S \rightarrow R$. Multiplicative terms (*) means logical AND, and additive terms means logical OR. Inclusion and coverage for sufficiency (inclS, covS), the unique coverage (covU), and the proportional reduction in inconsistency (PRI) allow comparing the different contributions of each term to the solutions.}
\label{fig:sols}
\end{figure*}
```

\pagebreak



# Supplementary Material {#sec:SM}

```{=tex}
\renewcommand\thefigure{A\arabic{figure}}
\renewcommand\thetable{A\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{sm1_correlogram}
\caption{\textbf{Correlogram} Matrix of correlations between features of adaptive capacity before calibration.}
\label{fig:sm1}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{heatmap_2tier}
\caption{\textbf{Adaptive capacity} Features at the second tier proposed by Berkes et al with raw scores before calibration.}
\label{fig:sm2}
\end{figure*}
```
